---
title: "Introduction to intraitR"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to intraitR}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 6,
  fig.height = 5
)
```

```{r setup}
library(intraitR)
```

## Overview

`intraitR` supports a complete workflow for the analysis of morphological
traits in freshwater fish, from digitized landmark coordinates to
morphological ratios, morphological space, intraspecific variability, and
measurement error. This vignette walks through the workflow end to end
using a simulated data set, so that it runs without any real specimen
data.

The statistical core of the package (Generalised Procrustes Analysis,
shape principal component analysis, Procrustes ANOVA and morphological
disparity) is delegated to the `geomorph` package. `intraitR` adds
fish-specific conveniences on top: file import, linear ratios, and
reporting functions geared towards ecomorphological questions.

## 1. Obtaining landmark data

Landmark data can be imported from a `tpsDig`-formatted `.tps` file with
[read_tps()], or from a generic long-format CSV/data.frame with
[read_landmarks_csv()]. For this vignette we instead simulate a data set
of three artificial species, each digitized three times per individual,
using [simulate_fish_landmarks()]:

```{r}
set.seed(1)
fish <- simulate_fish_landmarks(
  n_per_species = 15,
  species = c("Species_A", "Species_B", "Species_C"),
  n_landmarks = 12,
  n_replicates = 3
)
fish
head(fish$metadata)
```

Every specimen in `fish$metadata` is tagged with its species, population,
simulated standard length, and digitization replicate number.

A single configuration can be inspected visually with [plot_landmarks()]:

```{r}
plot_landmarks(fish, specimen = 1)
```

When the original specimen photographs are available, `plot_landmarks()`
(and [plot_fishmorph_points()], below) can overlay the landmarks directly
on the photograph via `background_image`, a useful visual check that no
landmark was placed off the body outline:

```{r, eval = FALSE}
plot_landmarks(fish, specimen = 1, background_image = "specimen1.jpg")
```

If landmarks have not yet been digitized, [digitize_landmarks()] provides
an interactive, point-and-click front end (built on
[geomorph::digitize2d()]) that digitizes directly from specimen
photographs and returns a ready-to-use `"intrait_landmarks"` object; it
requires a live interactive graphics session and so cannot be run inside
this vignette, but the intended usage is:

```{r, eval = FALSE}
lm <- digitize_landmarks(
  images = c("specimen1.jpg", "specimen2.jpg", "specimen3.jpg"),
  scheme = "fishmorph", tpsfile = "specimens.tps"
)
```

## 2. Generalised Procrustes Analysis

[gpa_fish()] superimposes all configurations, removing differences in
position, orientation and scale:

```{r}
gpa <- gpa_fish(fish)
gpa
```

`gpa$Csize` holds centroid size, the standard size measure retained
alongside Procrustes shape coordinates.

## 3. Linear distances and morphological ratios

Classical fish morphometric traits are expressed as ratios of linear
inter-landmark distances to a reference distance (typically standard
length). Distance pairs are defined by landmark index and computed on the
**raw** (pre-Procrustes) coordinates, since Procrustes superimposition
removes size:

```{r}
distances <- list(SL = c(1, 7), BD = c(3, 10), HL = c(1, 4))
morpho_ratios(fish, distances, norm_by = "SL")[1:5, ]
```

## 4. Morphological space

A morphospace is built from a Principal Component Analysis of the
Procrustes shape coordinates. By default, `plot()` displays each group as
its individual points, a "spider" of dashed segments linking every point
to its group mean, the group mean itself, and a 95% dispersion ellipse
(`style = "spider"`); pass `style = "hull"` for a classical convex-hull
display, `style = "density"` for a non-parametric kernel-density contour
(useful when a group's points are visibly skewed or multimodal, so a
bivariate-normal ellipse would be a poor fit), or `style = "none"` for
plain points. The group legend is drawn just outside the plot box by
default (`legend_position = "outside"`), so it never overlaps the data;
pass a standard `legend()` keyword (e.g. `"topright"`) to draw it inside
the plot box instead, as in previous versions. Axis limits are computed
to fully contain the ellipse/hull/contour, not just the raw points, so
none of these are ever clipped at the plot box edge. When `groups`
represents species identity, `legend_title`, `legend_italic`, and
`abbreviate_species` produce a properly typeset "Species" legend
(e.g. `"Barbatula barbatula"` displayed as italic *B. barbatula*):

```{r}
ms <- morpho_space(gpa, groups = fish$metadata$species)
ms
plot(ms, legend_title = "Species", legend_italic = TRUE, abbreviate_species = TRUE)
plot(ms, style = "hull", legend_title = "Species", legend_italic = TRUE,
     abbreviate_species = TRUE)
plot(ms, style = "density", legend_title = "Species", legend_italic = TRUE,
     abbreviate_species = TRUE)
```

## 5. Intraspecific morphological variability

[intraspecific_variability()] reports both a shape-based disparity
measure (Procrustes variance per group, with a permutation test) and
classical coefficients of variation for chosen linear traits:

```{r}
ratios <- morpho_ratios(fish, distances, norm_by = "SL")
iv <- intraspecific_variability(
  gpa = gpa,
  groups = fish$metadata$species,
  traits = ratios[, c("BD_ratio", "HL_ratio")],
  iter = 199
)
iv
```

[itv_index()] complements this with a formal variance-partitioning
approach (Violle et al., 2012; de Bello et al., 2011): the percentage of
each trait's total variance that lies *within* species (intraspecific
trait variability, ITV) versus *between* species (interspecific):

```{r}
itv <- itv_index(ratios[, c("BD_ratio", "HL_ratio")], groups = fish$metadata$species)
itv
plot(itv)
```

Because `fish$metadata` also records a `population` nested within each
species, the ITV component can be split further into a between-population
and a within-population (residual) part, following the
within-/among-population distinction commonly used in ITV meta-analyses
(e.g. Siefert et al., 2015):

```{r}
itv_nested <- itv_index(
  ratios[, c("BD_ratio", "HL_ratio")],
  groups = fish$metadata$species,
  nested = fish$metadata$population
)
itv_nested
plot(itv_nested)
```

## 6. Measurement error

Since our simulated data set includes three digitization replicates per
individual, [measurement_error()] can quantify how much of the observed
variance is attributable to digitization noise rather than genuine
biological variation, following a Procrustes ANOVA (Fruciano, 2016):

```{r}
individual_id <- sub("_rep[0-9]+$", "", fish$metadata$specimen)

me <- measurement_error(gpa, individual = individual_id, method = "procrustes", iter = 199)
me
```

For a single univariate trait (e.g. body depth ratio measured on the same
individuals multiple times), the classical ANOVA-based percent
measurement error and repeatability (Bailey & Byrnes, 1990) are also
available:

```{r}
me_ratio <- measurement_error(
  data.frame(value = ratios$BD_ratio, individual = individual_id),
  individual = "individual",
  method = "anova"
)
me_ratio
```

## 7. Allometry correction (optional)

If species differ substantially in size, [correct_allometry()] removes
the shape variation linearly associated with log centroid size before
further analysis:

```{r}
corrected <- correct_allometry(gpa, method = "group", groups = fish$metadata$species)
dim(corrected)
```

## 8. The FISHMORPH protocol (Brosse et al., 2021)

`intraitR` also implements the specific digitization and trait protocol
used to build the FISHMORPH database (Brosse et al., 2021), based on a
fixed scheme of 21 landmarks per specimen (snout, caudal fin basis, body
depth, head depth, eye position and diameter, mouth, pectoral fin, caudal
peduncle depth, caudal fin depth) plus an embedded scale bar, and
optionally a 22nd point to correct standard length for body curvature in
the picture.

We simulate a small data set following this exact scheme with
[simulate_fishmorph_points()], and inspect the digitization of one
specimen with [plot_fishmorph_points()]:

```{r}
fishmorph_fish <- simulate_fishmorph_points(n_per_species = 15, n_replicates = 1)
plot_fishmorph_points(fishmorph_fish, specimen = 1)
```

[fishmorph_segments()] computes the 11 linear measurements of the
protocol directly from the digitized points, automatically converting
pixel units to centimetres using the embedded scale bar (landmarks
20-21):

```{r}
segments <- fishmorph_segments(fishmorph_fish)
head(segments)
```

[fishmorph_ratios()] then derives the 9 unitless FISHMORPH ratios from
these measurements. The `no_caudal_fin`, `ventral_mouth`, and
`no_pectoral_fin` arguments implement the special-case rules of Villéger
et al. (2010) for species with unusual morphologies (e.g. no visible
caudal fin, a ventrally positioned mouth, or no pectoral fins):

```{r}
fishmorph_traits <- fishmorph_ratios(segments)
head(fishmorph_traits)
```

Finally, [trait_space()] builds a functional trait space from any numeric
trait table — here the 9 FISHMORPH ratios — by Principal Component
Analysis (or Principal Coordinate Analysis, `method = "pcoa"`). By
default, traits are `log10(x + 1)`-transformed (the `+ 1` accommodating
ratios that can legitimately equal 0 under the Villéger et al., 2010,
exception rules) and then centred and scaled to unit variance before the
ordination (`log_transform = TRUE`, `scale = TRUE`); set either to
`FALSE` to skip that step. As with `morpho_space()`, `plot()` defaults to
a spider/ellipse display per group:

```{r}
fts <- trait_space(fishmorph_traits, groups = fishmorph_fish$metadata$species)
fts
plot(fts)
```

Missing trait values, common in real ecomorphological data sets, are
handled via the `na_action` argument: `"fail"` (default) errors, `"omit"`
drops incomplete rows, `"impute_mean"`/`"impute_group_mean"` fill missing
values with column or within-group means, and `"missforest"` uses
nonparametric random-forest imputation (`missForest::missForest()`,
Stekhoven & Bühlmann, 2012), exploiting correlations among traits (and
`groups`, as an auxiliary predictor) rather than filling every missing
value in a trait with the same mean:

```{r, eval = FALSE}
fts_imputed <- trait_space(
  fishmorph_traits_with_gaps, groups = fishmorph_fish$metadata$species,
  na_action = "missforest"
)
```

Each option reports via `message()` how many rows or values were
affected (and, for `"missforest"`, the out-of-bag imputation error).

[trait_disparity()] complements `trait_space()` with a permutation test
of whether groups differ in multivariate trait dispersion (the trace of
each group's trait covariance matrix), computed on the full standardised
trait matrix rather than on the two axes shown in the plot above:

```{r}
td <- trait_disparity(fts, iter = 199)
td
```

[bootstrap_functional_space()] asks a related but different question:
does representing species by real individuals, rather than by their
centroid, inflate the estimated functional space (an n-dimensional
convex-hull volume in PCA space)? For each of `n_boot` bootstrap
"communities" it draws one individual at random per species and computes
the resulting convex-hull volume, and compares this distribution to a
single centroid-based reference volume, following Bertrand (2026);
requires the `geometry` package (Suggested, not installed by default):

```{r, eval = requireNamespace("geometry", quietly = TRUE)}
bf <- bootstrap_functional_space(fts, n_axes = 2, n_boot = 100)
bf
plot(bf)
```

[species_sensitivity()] then asks *which* species drive that difference:
each species' centroid is replaced, one individual at a time, by that
individual's own position (every other species fixed at its centroid),
and the resulting percent change in functional richness is summarised
per species as a mean effect and a min-max range (Bertrand, 2026):

```{r, eval = requireNamespace("geometry", quietly = TRUE)}
ss <- species_sensitivity(fts, n_axes = 2)
ss
plot(ss)
```

Both functions accept a `method` argument beyond the default
`"convexhull"`: `"dendrogram"` (UPGMA functional dendrogram branch
length, Petchey & Gaston 2002, no extra package required), `"tpd"`
(Trait Probability Density, Carmona et al. 2019), or `"hypervolume"`
(Gaussian-kernel hypervolume, Blonder et al. 2014, 2018). Since these
four measures are not on the same scale, [compare_functional_richness()]
runs several of them on the same data and tabulates the percent change
and bootstrap p-value side by side, so agreement (or disagreement) across
methods can itself be assessed:

```{r}
cmp <- compare_functional_richness(fts, methods = c("dendrogram", "convexhull"), n_axes = 2, n_boot = 100)
cmp
plot(cmp)
```

Because `fishmorph_segments()` and `fishmorph_ratios()` return plain
data frames, they compose directly with [summary_traits()] and
[intraspecific_variability()] (traits-only mode) exactly as the ratios
from [morpho_ratios()] do earlier in this vignette.

## References

Adams DC, Collyer ML, Kaliontzopoulou A (2024). geomorph: Software for
geometric morphometric analyses. R package.

Bailey RC, Byrnes J (1990). A new, old method for assessing measurement
error in both univariate and multivariate morphometric studies. Systematic
Zoology, 39(2), 124-130.

Brosse S, Charpin N, Su G, Toussaint A, Herrera-R GA, Tedesco PA, Villéger
S (2021). FISHMORPH: A global database on morphological traits of
freshwater fishes. Global Ecology and Biogeography, 30(11), 2330-2336.

de Bello F, Lavorel S, Albert CH, Thuiller W, Grigulis K, Dolezal J,
Janecek S, Leps J (2011). Quantifying the relative importance of
intraspecific trait variability and interspecific trait turnover for
functional diversity. Methods in Ecology and Evolution, 2(2), 163-174.

Fruciano C (2016). Measurement error in geometric morphometrics.
Development Genes and Evolution, 226, 139-158.

Siefert A, Violle C, Chalmandrier L, et al. (2015). A global meta-analysis
of the relative extent of intraspecific trait variation in plant
communities. Ecology Letters, 18(12), 1406-1419.

Stekhoven DJ, Bühlmann P (2012). MissForest -- non-parametric missing
value imputation for mixed-type data. Bioinformatics, 28(1), 112-118.

Villéger S, Ramos Miranda J, Flores Hernandez D, Mouillot D (2010).
Contrasting changes in taxonomic vs. functional diversity of tropical
fish communities after habitat degradation. Ecological Applications,
20(6), 1512-1522.

Violle C, Enquist BJ, McGill BJ, Jiang L, Albert CH, Hulshof C, Jung V,
Messier J (2012). The return of the variance: intraspecific variability
in community ecology. Trends in Ecology & Evolution, 27(4), 244-252.

Winemiller KO (1991). Ecomorphological diversification in lowland
freshwater fish assemblages from five biotic regions. Ecological
Monographs, 61(4), 343-365.

Zelditch ML, Swiderski DL, Sheets HD (2012). Geometric Morphometrics for
Biologists: A Primer (2nd ed). Academic Press.
